Comparing ChatGPT-4 to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6942 Deaths in Sierra Leone from 2019-2022

### 2024-01-24 manuscript/manuscript.tex
Edit methods

\subsection{Cause of Death Prediction Models}\label{methods-models}

Four computer models were used to predict COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. After conversion of all model and physician outputs to generalized Centre for Global Health Research 10 (CGHR-10) codes (see Additional File 2), 6939 records were comparable to evaluate performances of models for COD assignment relative to physicians.

@@

\subsubsection{Input Data}\label{methods-models-input}

For ChatGPT-3.5 and ChatGPT-4, text prompts were generated for each record as input to instruct the models to assign CODs based on the open narratives. These prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

\noindent For InterVA-5 and InSilicoVA, the standardized questionnaire data from the HEAL-SL EVA were first converted into 2016 World Health Organization (WHO) VA questionnaire revision 1.5.1 Open Data Kit (ODK) format \cite{nafundiODKCollectData2023}, followed by further conversion into OpenVA format \cite{liOpenVAToolkitVerbal2023} using the \verb|pyCrossVA| Python package \cite{thomasPycrossvaPrepareData}, and used as input for InterVA-5 and InSilicoVA. The 6942 physician agreed records were all converted into generated text prompts as ChatGPT-3.5 and ChatGPT-4 inputs. Only 6897 of the 6942 physician agreed records were converted into OpenVA formatted records as InterVA-5 and InSilicoVA inputs, where 45 records caused conversion errors and were skipped.

\subsubsection{Output Data}\label{methods-models-output}

@@

\subsection{Performance Evaluation}\label{methods-eval}

@@

CSMF accuracy measures the performance of models at the population level, comparing distributions of CODs between the physicians and the models \cite{murrayRobustMetricsAssessing2011}. To calculate CSMF accuracy, we first calculate $CSMF_j$ as is the fraction of physician or model records for cause $j$, given by dividing the number of records for cause $j$ with the total number of records as seen in Equation \ref{eq-csmf}. Then, the $CSMFMaximumError$, representing the worst possible model, is calculated using Equation \ref{eq-csmferr}. Finally, the CSMF accuracy is given by Equation \ref{eq-csmfa}, where $k$ is the number of causes, $j$ is a cause, $CSMF^{true}_j$ is the true physician CSMF for cause $j$, and $CSMF^{pred}_j$ is the predicted model CSMF for cause $j$. CSMF accuracy ranges from 0 to 1, where 1 means that the model completely matched the physician COD distribution and 0 means that it did not match the distribution at all.

>>>>

\subsection{Models for COD Assignment}\label{methods-models}

Four computer models were used to assign COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. After conversion of all model and physician outputs to generalized Centre for Global Health Research 10 (CGHR-10) codes (see Additional File 2), 6939 records were comparable to evaluate performances of models for COD assignment relative to physicians.

@@

\subsubsection{Input Data and Model Parameters}\label{methods-models-input}

For ChatGPT-3.5 and ChatGPT-4, text prompts were generated for each record as input to instruct the models to assign CODs based on the open narratives. Parameter \verb|t_prob| was set to 0.3. Prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

For InterVA-5 and InSilicoVA, the standardized questionnaire data from the HEAL-SL EVA were first converted into 2016 World Health Organization (WHO) VA questionnaire revision 1.5.1 Open Data Kit (ODK) format \cite{nafundiODKCollectData2023}, followed by further conversion into OpenVA format \cite{liOpenVAToolkitVerbal2023} using the \verb|pyCrossVA| Python package \cite{thomasPycrossvaPrepareData}, and used as input for InterVA-5 and InSilicoVA. The \verb|Nsim| (number of iterations to run) parameter for InSilicoVA was set to 10,000, while the \verb|HIV| (level of prevalence of human immunodeficiency virus) and \verb|Malaria| (level of prevalence of Malaria) parameters were both set to \verb|'h'| (high).

The 6942 physician agreed records were all converted into instructional text prompts for ChatGPT-3.5 and ChatGPT-4. Only 6897 of the 6942 physician agreed records were converted into OpenVA formatted records for InterVA-5 and InSilicoVA, where 45 records caused conversion errors and were skipped or removed.

\subsubsection{Output Data with CGHR-10 Codes}\label{methods-models-output}

@@

\subsection{Comparing Models to Physician COD Assignment}\label{methods-eval}

@@

CSMF accuracy measures the performance of models at the population level, comparing distributions of CODs between the physicians and the models \cite{murrayRobustMetricsAssessing2011}. To calculate CSMF accuracy, we first calculate $CSMF_j$ as is the fraction of physician or model records for cause $j$, given by dividing the number of records for cause $j$ with the total number of records as seen in Equation \ref{eq-csmf}. Then, the $CSMFMaximumError$, representing the worst possible model, is calculated using Equation \ref{eq-csmferr}. Finally, the CSMF accuracy is given by Equation \ref{eq-csmfa}, where $k$ is the number of causes, $j$ is a cause, $CSMF^{true}_j$ is the true physician CSMF for cause $j$, and $CSMF^{pred}_j$ is the prediction model CSMF for cause $j$. CSMF accuracy ranges from 0 to 1, where 1 means that the model completely matched the physician COD distribution and 0 means that it did not match the distribution at all.
